Mapping the current knowledge in syndemic research applied to Men who have Sex with Men: a Scoping Review

### 2022-04-22 _bookdown_files/01-abstract.Rmd
Reviews [1st round]

Men who have sex with Men (MSM) represent a vulnerable population affected by numerous health conditions. Syndemic theory has been used as a framework to study the health of MSM for nearly 20 years. However, the literature is plagued by a lack of consensus regarding what constitutes a synergy in a syndemic and recent reviews have shown that most of the papers published thus far failed to demonstrate a synergy nor describe the bio-social interaction needed to account for a true syndemic. Moreover, to our knowledge, none of the existing reviews have focused specifically on MSM.

This scoping review aims to fill this gap by mapping in details how syndemic research on MSM was conducted. A systematic database search was conducted between 2020 and 2021 and 115 studies were included. Our findings showed a lack of diversity regarding the location, design, subpopulation and outcomes studied. In addition, the syndemic conditions as well as their measurement were not focused enough to ensure the robustness and reproducibility of the findings. Furthermore, our results support previous reviews showing a lack of empirical data to support disease interaction in syndemic research applied to MSM. Our review offers some important recommendations to help move the field forward in future work and describes some promising methodological advancement.

>>>>

Men who have sex with Men (MSM) represent a population affected by numerous health conditions. Syndemic theory has been used as a framework to study the health of MSM for nearly 20 years. However, the literature is plagued by a lack of consensus regarding what constitutes a synergy in a syndemic and recent reviews have shown that most of the papers published thus far have failed to demonstrate a synergy nor describe the bio-social interaction needed to account for a true syndemic. Moreover, to our knowledge, none of the existing reviews have focused specifically on MSM.

This scoping review aims to fill this gap by mapping in detail how syndemic research on MSM has been conducted. A systematic database search was conducted between 2020 and 2021 and 115 studies were included. Our findings showed a lack of diversity regarding the location, design, subpopulation and outcomes studied. In addition, the syndemic conditions as well as their measurement were not focused enough to ensure the robustness and reproducibility of the findings. Furthermore, our results support previous reviews showing a lack of empirical data to support disease interaction in syndemic research applied to MSM. Our review offers some important recommendations to help move the field forward in future work and describes some promising methodological advances.

### 2022-04-22 _bookdown_files/02-introduction.Rmd
Reviews [1st round]

Men who have sex with men (MSM) represent a vulnerable population disproportionately affected by numerous health conditions. They represent 70% of new HIV diagnoses [@centerfordiseasecontrolandprevention2020] in the United States of America and more than half of new HIV diagnoses in European Union/European Economic area [@europeancentrefordiseasepreventionandcontrol2019]. They are also more prone to contract other sexually transmitted diseases and *Nesseria Gonorrhea* strains from MSM exhibit higher antimicrobial resistance [@centersfordiseasecontrolandprevention2019]. Furthermore, mental health conditions such as depression, anxiety, suicide attempts or self-harm are more prevalent among MSM [@ross2018; @Liu2019; @luo2017] and substance use is more common than for their heterosexual counterparts[@Medley2016]. Moreover, those adverse health outcomes are enmeshed within structural disadvantages such as violence, stigma, discrimination [@collier2013; @Lee2016; @Lea2014] as well as with poverty, unemployment, unstable housing and poor healthcare access [@closson2018; @ayhan2020].

For more than 20 years, syndemic theory has provided a framework to examine the interrelations between social conditions, mental health and physical health [@singer2017]. Described for the first time in 1996 by Merrill Singer [@singer1996], a syndemic consists of two more interacting epidemics producing an excess burden in a population because of harmful social conditions [@singer2003]To speak of a syndemic, three conditions must be met [@mendenhall2020] :

1.  Two or more conditions cluster in a given population

2.  This clustering is due to an adverse social context such as poverty or stigmatization

3.  There is some form of biological, social and/or behavioral interaction between the conditions, significantly worsening the health of the affected population

Syndemic is thus an holistic framework, describing interactions both between diseases themselves and between diseases and the social environment contributing to their emergence, clustering and spread [@singer2017]. As such, its value for studying the health of marginalized populations such as MSM is undeniable.

@@

However this framework is frequently misused by researchers. A systematic review conducted in 2015 showed the inadequacy of most of the literature in supporting one of its core tenets [@tsai2015]. Indeed, this paper demonstrated that, although synergistic interaction between diseases lies in the core of the theory, most papers failed to use relevant statistics to demonstrate the existence of a synergy. More recently, a scoping review published in 2020 confirmed this finding as most recently-published citations failed to describe the interactions between diseases needed to account for a true syndemic [@Singer2020]. Scholars in the field have been calling for greater clarity in the use of the concept and urged researchers to describe more precisely the interactions they observe between health conditions [@singer2021 ; @mendenhall2020].

However, to our knowledge none of the existing reviews or recommendations have focused on syndemic literature applied specifically on MSM. We thus sought to fill this gap by conducting a scoping review with the objective to map the following :

1.  the study design used ;

2.  the subpopulations of MSM studied ;

3.  the psychosocial conditions considered as forming a syndemic and how they were measured;

4.  the outcomes studied;

5.  the statistics used to evaluate the concept of interaction;

6.  the hypothesis for biological and sociobiological interactions proposed by the authors, when available;

7.  relevant frameworks that may compliment syndemic theory to better understand the health of MSM and

8.  the key findings of these studies

>>>>

Men who have sex with men (MSM) represent a population disproportionately affected by numerous health conditions. They represent 70% of new HIV diagnoses [@centerfordiseasecontrolandprevention2020] in the United States of America (USA) and more than half of new HIV diagnoses in the European Union/European Economic area [@europeancentrefordiseasepreventionandcontrol2019]. They are also more prone to contract other sexually transmitted diseases and *Nesseria Gonorrhoea* strains from MSM exhibit higher antimicrobial resistance [@centersfordiseasecontrolandprevention2019]. Furthermore, mental health conditions such as depression, anxiety, suicide attempts or self-harm are more prevalent among MSM [@ross2018; @Liu2019; @luo2017] and substance use is more common than for their heterosexual counterparts [@Medley2016]. Moreover, those adverse health outcomes are enmeshed within structural disadvantages such as violence, stigma, discrimination [@collier2013; @Lee2016; @Lea2014] as well as with poverty, unemployment, unstable housing and poor access to healthcare [@closson2018; @ayhan2020].

For more than 20 years, syndemic theory has provided a framework to examine the interrelations between social conditions, mental health and physical health [@singer2017]. Described for the first time in 1996 by Merrill Singer [@singer1996], a syndemic consists of two or more interacting epidemics producing an excess burden in a population due to harmful social conditions [@singer2003]. To speak of a syndemic, three conditions must be met [@mendenhall2020] :

1.  Two or more conditions cluster in a given population.

2.  This clustering is due to an adverse social context such as poverty or stigmatisation.

3.  There is some form of biological, social and/or behavioural interaction between the conditions, significantly worsening the health of the affected population.

A syndemic is thus a holistic framework describing interactions both between diseases themselves and between diseases and the social environment contributing to their emergence, clustering and spread [@singer2017]. As such, its value for studying the health of marginalised populations such as MSM is undeniable.

@@

Nevertheless, this framework is frequently misused by researchers. A systematic review conducted in 2015 showed the inadequacy of most of the literature in supporting one of its core tenets [@tsai2015]. Indeed, that paper demonstrated that, although synergistic interaction between diseases lies at the core of the theory, most papers failed to use relevant statistics to demonstrate the existence of a synergy. More recently, a scoping review published in 2020 confirmed this finding that most recently published citations did not describe the interactions between diseases needed to account for a true syndemic [@Singer2020]. Scholars in the field have been calling for greater clarity in the use of the concept and have urged researchers to describe more precisely the interactions they observe between health conditions [@singer2021 ; @mendenhall2020].

However, to our knowledge none of the existing reviews or recommendations have focused on syndemic literature applied specifically to MSM. We thus sought to fill this gap by conducting a scoping review with the objective to map the following:

1.  the study design used,

2.  the subpopulations of MSM studied,

3.  the psychosocial conditions considered as forming a syndemic and how they were measured,

4.  the outcomes studied,

5.  the statistics used to evaluate the concept of interaction,

6.  the hypothesis for biological and sociobiological interactions proposed by the authors, when available,

7.  relevant frameworks that may compliment syndemic theory to better understand the health of MSM and,

8.  the key findings of these studies.

### 2022-04-22 _bookdown_files/03-methodology.Rmd
Reviews [1st round]

# Methodology

The complete protocol of this scoping review was published in 2020 [Anonymous 2020].

We applied the framework suggested by Arskey and O'Malley [@arksey2005], with enhancements from Levac et al. [@levac2010]. We also took into account the recommendations formulated by Colqhoun et al. [@colquhoun2014] and the updated guidance by Peters and colleagues [@peters2020]. We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses extension for Scoping Reviews (PRISMA-ScR) [@Tricco2018]. The completed checklist can be found in \@ref(tab:PRISMA-ScR) of the supplementary materials.

Screening and data charting was completed using DistillerSR (Evidence Partners, Ottawa, Canada). Scripts, statistics, charts, tables and the present paper was generated using R [@rcoreteam2020], RStudio [@rstudioteam2020] and appropriate packages. To ensure the complete transparency and reproducibility of this article, our data manager (M.O.) uploaded every data used to generate this article on Open Science Framework (DOI : [masked for the anonymity of the review])

Our main research question : *"What is known about Syndemic Theory applied to MSM"* was subdivided into three sub-questions :

1.  *"How are studies concerning Syndemic Theory applied to MSM conducted ?"*;
2.  *"How is the concept of interaction explored in syndemic research applied to MSM ?"* and
3.  *"What were the key findings of these studies ?"*.

To answer these questions, we conducted a systematic search strategy in the following databases : Medline, PsycInfo, Scopus, Cochrane Central Register of Controlled Trials and ProQuest Sociological Abstracts using complex search equations to include every synonym for "men who have sex with men". The complete search strategy for every database can be found in the Supplementary Material of our protocol. No date limit was applied given the relative novelty of the syndemic literature. A first search using this strategy was conducted the 11th of June 2020 and was subsequently updated on the 9th of February 2021 and finally on the 11th of November 2021.

Search results were downloaded into .RIS files and imported into DistillerSR. After duplicate removal, we screened the title and abstracts for eligibility, using forms generated with DistillerSR by the authors. Articles that met the inclusion criteria or for which eligibility was unclear underwent a second screening in which the full texts were assessed for eligibility. As per the protocol, 10% of the titles and abstracts were screened by two reviewers. A Kappa inter-rater reliability score was computed using DistillerSR and was equal to 0.86. As it exceeded our cutoff of 0.8, the rest of the screening process was conducted by the main investigator only. Articles were included if they met the following criteria :

-   MSM was either the only population studied in the paper or, if studied alongside other populations, such as transgender women, disaggregated data must have been available for MSM

-   Syndemic framework was the main focus of the study

-   Studies were cohort, case-control, cross-sectional, controlled trials, mixed studies, qualitative studies, systematic reviews or meta-analyses. We excluded letters, commentaries, conference abstracts, editorials or narrative reviews

@@

After inclusion of relevant studies, we hand-searched the reference lists to manually add pertinent studies. The same selection process was then applied and these papers are marked as "Additional records identified through other sources" in the study selection flow diagram.

Data charting was also performed using forms generated in DistillerSR by the authors. The complete list of all variables for which data were sought can be found in our protocol [Anonymous 2020]. Furthermore, the README file of our *Data* subdirectory contains an exhaustive definition of every variable included. Two amendments were made to our initial protocol :

1.  We included a variable to chart any additional framework the authors might have used

2.  We included variables to chart data for systematic reviews and meta-analyses. These two kind of studies were omitted from our initial protocol as we were not aware of the existence of systematic reviews on syndemic theory applied to MSM by the time we published the protocol. The variables included were general characteristics (authors, year, location, design, total sample size), purpose of the review, number of studies included, hypothesis for biological or bio-social interaction, key findings and additional framework.

Data was then collated, summarised and reported using R, RStudio and relevant packages. Charts were generated to better visualize year of publication, location, population and syndemic conditions studied. Tables were generated to summarise the main variables of every article as well as the measurement methods of the most studied syndemic conditions. These tables can be found respectively in part B and C of the supplementary materials of this paper. Additionally, we generated an online report accessible at <https://datastudio.google.com/u/0/reporting/56c1e17a-fa3f-47e9-9516-d3d0eef7ec1> This report contains every study included and provide scholars in the field the opportunity to search relevant studies using multiple filters.

>>>>

# Materials and Methods

The complete protocol of this scoping review was published in 2020 [Author (2020 of publication) Details omitted for double-blind review.].

We applied the framework suggested by Arskey and O'Malley [@arksey2005], with enhancements from Levac et al. [@levac2010]. We also considered the recommendations formulated by Colquhoun et al. [@colquhoun2014] and the updated guidance by Peters and colleagues [@peters2020]. We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses extension for Scoping Reviews (PRISMA-ScR) [@Tricco2018]. The completed checklist can be found in Table \@ref(tab:PRISMA-ScR) of the supplementary materials.

Screening and data charting were completed using DistillerSR (Evidence Partners, Ottawa, Canada). Scripts, statistics, charts, tables and the present paper was generated using R [@rcoreteam2020], RStudio [@rstudioteam2020] and the appropriate packages. To ensure the complete transparency and reproducibility of this article, the lead author (M.O.) uploaded all data used to generate this article on Open Science Framework ([[DOI:[masked](DOI:%5Bmasked](DOI:%5Bmasked%5D(DOI:%5Bmasked)%7B.uri%7D){.uri} for the anonymity of the review])

Our main research question was: *"What is known about Syndemic Theory applied to MSM"* was subdivided into three sub-questions:

1.  *"How are studies concerning Syndemic Theory applied to MSM conducted?"*;
2.  *"How is the concept of interaction explored in syndemic research applied to MSM?"* and
3.  *"What were the key findings of these studies?"*

To answer these questions, we conducted a systematic search strategy in the following databases: Medline, PsycInfo, Scopus, Cochrane Central Register of Controlled Trials and ProQuest Sociological Abstracts using complex search equations to include every synonym for "men who have sex with men." The complete search strategy for each database can be found in the Supplementary Material of our protocol. No date limit was applied given the relative novelty of the syndemic literature. A first search using this strategy was conducted on the 11th of June 2020 and was subsequently updated on the 9th of February 2021 and finally on the 11th of November 2021.

Search results were downloaded into .RIS files and imported into DistillerSR. After duplicate removal, we screened the title and abstracts for eligibility, using forms generated with DistillerSR by the authors. Articles that met the inclusion criteria or for which eligibility was unclear underwent a second screening in which the full texts were assessed for eligibility. As per the protocol, 10% of the titles and abstracts were screened by two reviewers. A Kappa inter-rater reliability score was computed using DistillerSR and was equal to 0.86. As it exceeded our cutoff of 0.8, the rest of the screening process was conducted by the main investigator only. Articles were included if they met the following criteria:

-   MSM was either the only population studied in the paper or, if studied alongside other populations, such as transgender women, disaggregated data must have been available for MSM.

-   The syndemic framework was the main focus of the study.

-   Studies were cohort, case-control, cross-sectional, controlled trials, mixed studies, qualitative studies, systematic reviews or meta-analyses. We excluded letters, commentaries, conference abstracts, editorials or narrative reviews.

@@

After inclusion of relevant studies, we hand-searched the reference lists to manually add pertinent studies. The same selection process was then applied, and these papers are marked as "Additional records identified through other sources" in the study selection flow diagram (Figure \@ref(fig:PRISMA)).

Data charting was also performed using forms generated in DistillerSR by the authors. The complete list of all variables for which data were sought can be found in our protocol[Anonymous 2020]. Furthermore, the README file of our *Data* subdirectory contains an exhaustive definition of every variable included. Two amendments were made to our initial protocol:

1.  We included a variable to chart any additional frameworks the authors might have used.

2.  We included variables to chart data for systematic reviews and meta-analyses. These two kinds of studies were omitted from our initial protocol as we were not aware of the existence of systematic reviews on syndemic theory applied to MSM at the time we published the protocol. The variables included were general characteristics (authors, years, location, design, total sample size), purpose of the review, number of studies included, hypotheses for biological or bio-social interaction, key findings and additional frameworks.

Data were then collated, summarised and reported using R, RStudio and relevant packages. Charts were generated to better visualise years of publication, location, population and syndemic conditions studied. Tables were generated to summarise the main variables of every article as well as the measurement methods of the most studied syndemic conditions. These tables can be found respectively in part B and C of the supplementary materials of this paper. Additionally, we generated an online report accessible at (link omitted for double-blind review). This report contains all studies included in the review and provides researchers in the field the opportunity to search relevant studies using multiple filters.
